Dark matter web in the Universe

### 2018-06-12 Draft/Chapter2/background.tex
spell checked

The most fundamental attributes of particles in the N-body simulations are their position and velocity co-ordinates, and their masses. Due to the lack of numerical tools for direct analysis these raw data, fields such as mass density $\rho(\mathbf{x}, t)$, velocity $\mathbf{v}(\mathbf{x}, t)$ or gravitational potential $\phi(\mathbf{x}, t)$ fields are often computed numerically. Mass density fields were calculated using Cloud-in-Cell (CIC) algorithm (cf. \citealt{Hockney1988}), which is numerically equivalent to counting the number of particles on each cell of a regular grid. Alternatively, the density field also generated on irregular grids by applying Delaunay (For example, \citealt{Icke1991} and the Delaunay Tessellation Field Estimator (DTFE) by \citealt{Schaap2000} and \citealt{Weygaert2009a}) or Voronoi tessellations (See \citealt{Schaap2000} and references therein) to the particle coordinates. Another parameter `linking length', using distances between nearest neighbouring particles, was used for percolation analyses and identifying super-clusters of galaxies ( \citealt{Zeldovich1982}, \citealt{Shandarin1983} and \citealt{Shandarin1983b}) for identifying halos \citealt{Davis1985} in the cosmological simulations. Left panel in \autoref{fig:cosmicfields} shows some of the popular fields/parameters that use particle mass and positions. It has to be noted that the density fields or linking-lengths are not dynamical descriptions that invoke the initial field of density fluctuations or the velocity of the particles.

\begin{sidewaysfigure}
\begin{minipage}[t]{0.99\linewidth}
 \centering\includegraphics[height=10.5cm]{Chapter1/Plots/fig0.pdf} 
\end{minipage}\hfill
\captionsetup{font={small,stretch=1}}
\caption{Classification of some of the fields/parameters used in cosmological analyses. Some fields utilise position co-ordinates only whereas others use the full phase-space information. In addition, the fields may be defined on a regular grid, or may be defined on an unstructured grid (For instance, flip-flop is a number-valued field defined on each dark matter particle). Fields like mass density and multi-streams can be defined on either grids, depending on the numerical technique. List is obviously not exhaustive- velocity and potential fields are not included, and discussions of correlation functions are excluded as well.}
\label{fig:cosmicfields}
\end{sidewaysfigure}

An obvious advantage of methods based on particle coordinates, both on structured and unstructured grids, is their applicability to redshift catalogues. The redshift catalogues like SDSS and 2dF provide only two angular coordinates and distances in redshift space. But cosmological N-body dark matter simulations provide the full dynamical information in six-dimensional phase space. This additional information is very valuable providing a greater opportunity for understanding the physics of the web and developing a better theory of the web.

@@

ZA is an elegant analytical approximation to describe the non-linear gravitational evolution of collisionless particles in continuous media. Technically it is the first oder Lagrangian perturbation theory, however Zeldovich suggested to extrapolate it to the beginning of the non-perturbative nonlinear stage and predicted the formation of caustics which are the boundaries of
the first very thin multistream regions dubbed by him  'pancakes'.
ZA describes a dynamical mapping from the initial Lagrangian coordinates $\mathbf{q}$ to Eulerian positions at time $t$. In comoving coordinates, $\mathbf{x} = \mathbf{r}/a(t)$ (where $\mathbf{r}$ is the physical coordinate  and $a(t)$ is the scale factor; assuming normalization $a(z=0)=1, \mathbf{r}$ are 
the physical coordinates of particles at present ), ZA takes the form:

@@

Complexities in 3-dimensional caustics is partly due the intricate mapping in the hypersurface $\mathbf{x}(\mathbf{q})$ called the Lagrangian submanifold (See Figure \ref{fig:Tess}). The Lagrangian submanifold $\mathbf{x}(\mathbf{q})$  is a single valued, smooth and differentiable function in it's 6-dimensional space $(\mathbf{q}, \mathbf{x})$, however the projection onto 3-dimensional Eulerian space is entangled with creases, kinks and folds (Note that this  submanifold is very different than the phase space $(\mathbf{x},\mathbf{v})$, even though they are connected by a cannonical transformation). However, dilineating the Lagrangian submanifold reveals several properties of the dark matter dynamics not inferred from position-space analyses. Two fields related to tessellating the Lagrangian submanifold -- The Multistream field $n_{str}(\mathbf{x})$ in Eulerian space and the Flip-Flop field $n_{ff}(\mathbf{q})$ in Lagrangian space (check \cite{Shandarin2012}, \cite{Ramachandra2015}, \cite{Shandarin2016}) are closely related.

@@

The bottom panel \autoref{fig:phase1d} shows the the multistream field $n_{str}(\bf{x})$ at $z = 0$. The field only takes the values of 1, 3, 5 and 7 in this scenario. Caustics occur at the folds in Lagrangian sub-manifold, and have a measure zero (study of caustics in one- and  two-dimensional evolution is done in \cite{Hidding2014}, three-dimensional caustic surface in a cosmological simulation is shown in \cite{Ramachandra2017} ). Several properties of the multistream field are significantly different from mass density. The bottom panel also shows an illustration of CIC algorithm (cf. \citealt{Hockney1988}) in calculating density, which is numerically equivalent to counting the number of particles on each cell of a regular grid. One major difference is in the regions before gravitational collapse: $n_{str}$ is universally equal to unity, whereas number density fluctuates. It should also be noted that density by definition is a continuous field; numerical approximations like CIC discretise the field. Alternatively, multistream field is intrinsically a discrete-data field.

@@

\begin{figure} 
\centering\includegraphics[width=16cm]{Chapter4/Source_v2/fig1.pdf} 
\caption{3D rendering of the multistream field: the cosmic web structure of a $ 50 h^{-1} \text{Mpc} \times 50 h^{-1} \text{Mpc} \times 50 h^{-1} \text{Mpc}$ slice in a simulation box of side length $100 h^{-1}$ Mpc and $128^3$ particles. The multistream field is calculated at 8 times the native resolution. void(black) is a percolating structure with $n_{str} = 1$. Regions $n_{str} \geq 17$ show a filamentary structure (gray) and the bright spots at the intersections of the filaments are regions with $n_{str} \geq 100$. }
\label{fig:full_3d}
\end{figure}

@@

\begin{figure} 
\centering\includegraphics[width=11cm]{Chapter5/Source_v2/fig2.pdf} 
\caption{Multistream field contours: The multistream field is calculated at 8 times the native resolution. Each inner convex blobs (red) surround local multistream maxima inside. Surrounding outer shell(blue) is not convex throughout the surface, and the outermost gray multistream surface displays a filamental geometry.}
\label{fig:full}
\end{figure}

Caustic formations and mass accretion are also seen to occur more along the higher streams, which makes the haloes non-spherical, with the alignment generally determined by a complicated interplay  of the intensities of the streams from neighboring filamentary structures. Number of streams corresponding to the dark matter halo also has a local environment dependence. The three small haloes in \autoref{fig:full}, where the number of streams are higher than the neighbouring filaments, are aligned along three intersecting filaments. Halo environment studied in \cite{Ramachandra2015} show similar hierarchical variation in $n_{str}$ values. The halo environments are thus very complicated, and empirical thresholds (on multistream or density fields) may not account for all the haloes uniformly. Hence we use a local geometrical method to identify potential haloes in multistream fields.

>>>>

The most fundamental attributes of particles in the N-body simulations are their position and velocity co-ordinates, and their masses. Due to the lack of numerical tools for direct analysis these raw data, fields such as mass density $\rho(\mathbf{x}, t)$, velocity $\mathbf{v}(\mathbf{x}, t)$ or gravitational potential $\phi(\mathbf{x}, t)$ fields are often computed numerically. Mass density fields were calculated using Cloud-in-Cell (CIC) algorithm (cf. \citealt{Hockney1988}), which is numerically equivalent to counting the number of particles on each cell of a regular grid. Alternatively, the density field also generated on irregular grids by applying Delaunay (For example, \citealt{Icke1991} and the Delaunay Tessellation Field Estimator (DTFE) by \citealt{Schaap2000} and \citealt{Weygaert2009a}) or Voronoi tessellations (See \citealt{Schaap2000} and references therein) to the particle coordinates. Another parameter `linking length', using distances between nearest neighboring particles, was used for percolation analyses and identifying super-clusters of galaxies ( \citealt{Zeldovich1982}, \citealt{Shandarin1983} and \citealt{Shandarin1983b}) for identifying halos \citealt{Davis1985} in the cosmological simulations. Left panel in \autoref{fig:cosmicfields} shows some of the popular fields/parameters that use particle mass and positions. It has to be noted that the density fields or linking-lengths are not dynamical descriptions that invoke the initial field of density fluctuations or the velocity of the particles.

\begin{sidewaysfigure}
\begin{minipage}[t]{0.99\linewidth}
 \centering\includegraphics[height=10.5cm]{Chapter1/Plots/fig0.pdf} 
\end{minipage}\hfill
\captionsetup{font={small,stretch=1}}
\caption{Classification of some of the fields/parameters used in cosmological analyses. Some fields utilize position co-ordinates only whereas others use the full phase-space information. In addition, the fields may be defined on a regular grid, or may be defined on an unstructured grid (For instance, flip-flop is a number-valued field defined on each dark matter particle). Fields like mass density and multi-streams can be defined on either grids, depending on the numerical technique. List is obviously not exhaustive- velocity and potential fields are not included, and discussions of correlation functions are excluded as well.}
\label{fig:cosmicfields}
\end{sidewaysfigure}

An obvious advantage of methods based on particle coordinates, both on structured and unstructured grids, is their applicability to redshift catalogs. The redshift catalogs like SDSS and 2dF provide only two angular coordinates and distances in redshift space. But cosmological N-body dark matter simulations provide the full dynamical information in six-dimensional phase space. This additional information is very valuable providing a greater opportunity for understanding the physics of the web and developing a better theory of the web.

@@

ZA is an elegant analytical approximation to describe the non-linear gravitational evolution of collisionless particles in continuous media. Technically it is the first oder Lagrangian perturbation theory, however Zel'dovich suggested to extrapolate it to the beginning of the non-perturbative nonlinear stage and predicted the formation of caustics which are the boundaries of
the first very thin multistream regions dubbed by him  'pancakes'.
ZA describes a dynamical mapping from the initial Lagrangian coordinates $\mathbf{q}$ to Eulerian positions at time $t$. In co-moving coordinates, $\mathbf{x} = \mathbf{r}/a(t)$ (where $\mathbf{r}$ is the physical coordinate  and $a(t)$ is the scale factor; assuming normalization $a(z=0)=1, \mathbf{r}$ are 
the physical coordinates of particles at present ), ZA takes the form:

@@

Complexities in 3-dimensional caustics is partly due the intricate mapping in the hyper-surface $\mathbf{x}(\mathbf{q})$ called the Lagrangian sub-manifold (See Figure \ref{fig:Tess}). The Lagrangian sub-manifold $\mathbf{x}(\mathbf{q})$  is a single valued, smooth and differentiable function in it's 6-dimensional space $(\mathbf{q}, \mathbf{x})$, however the projection onto 3-dimensional Eulerian space is entangled with creases, kinks and folds (Note that this  sub-manifold is very different than the phase space $(\mathbf{x},\mathbf{v})$, even though they are connected by a canonical transformation). However, delineating the Lagrangian sub-manifold reveals several properties of the dark matter dynamics not inferred from position-space analyses. Two fields related to tessellating the Lagrangian sub-manifold -- The Multistream field $n_{str}(\mathbf{x})$ in Eulerian space and the Flip-Flop field $n_{ff}(\mathbf{q})$ in Lagrangian space (check \cite{Shandarin2012}, \cite{Ramachandra2015}, \cite{Shandarin2016}) are closely related.

@@

The bottom panel \autoref{fig:phase1d} shows the the multistream field $n_{str}(\bf{x})$ at $z = 0$. The field only takes the values of 1, 3, 5 and 7 in this scenario. Caustics occur at the folds in Lagrangian sub-manifold, and have a measure zero (study of caustics in one- and  two-dimensional evolution is done in \cite{Hidding2014}, three-dimensional caustic surface in a cosmological simulation is shown in \cite{Ramachandra2017} ). Several properties of the multistream field are significantly different from mass density. The bottom panel also shows an illustration of CIC algorithm (cf. \citealt{Hockney1988}) in calculating density, which is numerically equivalent to counting the number of particles on each cell of a regular grid. One major difference is in the regions before gravitational collapse: $n_{str}$ is universally equal to unity, whereas number density fluctuates. It should also be noted that density by definition is a continuous field; numerical approximations like CIC discretize the field. Alternatively, multistream field is intrinsically a discrete-data field.

@@

\begin{figure} 
\centering\includegraphics[width=16cm]{Chapter4/Source_v2/fig1.pdf} 
\caption{Three dimensional rendering of the multistream field: the cosmic web structure of a $ 50 h^{-1} \text{Mpc} \times 50 h^{-1} \text{Mpc} \times 50 h^{-1} \text{Mpc}$ slice in a simulation box of side length $100 h^{-1}$ Mpc and $128^3$ particles. The multistream field is calculated at 8 times the native resolution. void(black) is a percolating structure with $n_{str} = 1$. Regions $n_{str} \geq 17$ show a filamentary structure (gray) and the bright spots at the intersections of the filaments are regions with $n_{str} \geq 100$. }
\label{fig:full_3d}
\end{figure}

@@

\begin{figure} 
\centering\includegraphics[width=11cm]{Chapter5/Source_v2/fig2.pdf} 
\caption{Multistream field contours: The multistream field is calculated at 8 times the native resolution. Each inner convex blobs (red) surround local multistream maxima inside. Surrounding outer shell(blue) is not convex throughout the surface, and the outermost gray multistream surface displays a filamentary geometry.}
\label{fig:full}
\end{figure}

Caustic formations and mass accretion are also seen to occur more along the higher streams, which makes the haloes non-spherical, with the alignment generally determined by a complicated interplay  of the intensities of the streams from neighboring filamentary structures. Number of streams corresponding to the dark matter halo also has a local environment dependence. The three small haloes in \autoref{fig:full}, where the number of streams are higher than the neighboring filaments, are aligned along three intersecting filaments. Halo environment studied in \cite{Ramachandra2015} show similar hierarchical variation in $n_{str}$ values. The halo environments are thus very complicated, and empirical thresholds (on multistream or density fields) may not account for all the haloes uniformly. Hence we use a local geometrical method to identify potential haloes in multistream fields.
